Practice Problems
Problem 44.1
Let $X$ and Y be two random variables with joint density function

otherwise.
Find $f_{X|Y}(x|y)$, the conditional probability density function of $X$ given Y =
y. Sketch the graph of f X|Y (x|0.5).



Problem 44.2
Suppose that $X$ and $Y$ have joint density function

8xy 0 ≤ $X$ < y ≤ 1
f_{XY}(x,y) =
0
otherwise.
Find $f_{X|Y}(x|y)$, the conditional probability density function of $X$ given Y =
y.




Problem 44.3
Suppose that $X$ and Y have joint density function
 3y 2
0 ≤ y<x ≤ 1
x 3
f_{XY}(x,y) =
0
otherwise.
Find f Y |X (y|x), the conditional probability density function of $Y$ given $X$ =
x.




Problem 44.4
The joint density function of $X$ and $Y$ is given by
 −x(y+1)
xe
x ≥ 0, y ≥ 0
f_{XY}(x,y) =
0
otherwise.
Find the conditional density of $X$ given $Y = y$ and that of $Y$ given $X$ = x.



Problem 44.5
Let $X$ and $Y$ be continuous random variables with conditional and marginal
p.d.f.’s given by

where I A (x) is the indicator function of A.
(a) Find the joint p.d.f. of $X$ and $Y$.
(b) Find the conditional p.d.f. of $X$ given $Y = y$.
f Y |X (y|x) =


Problem 44.6
Suppose $X$, $Y$ are two continuous random variables with joint probability
density function

12xy(1 − x) 0 < x, y < 1
f_{XY}(x,y) =
0
otherwise.
(a) Find $f_{X|Y}(x|y)$. Are $X$ and $Y$ independent?
(b) Find P(Y < 2 1 |X > 2 1 ).



Problem 44.7
The joint probability density function of the random variables $X$ and $Y$ is
given by
 1
x − y + 1 1 ≤ $X$ ≤ 2, 0 ≤ y ≤ 1
3
f_{XY}(x,y) =
0
otherwise.
(a) Find the conditional probability density function of $X$ given $Y$ = y.
(b) Find P(X < 32 |Y = 2 1 ).



Problem 44.8 ‡

Let $X$ and $Y$ be continuous random variables with joint density function

24xy 0 < $X$ < 1, 0 < y < 1 − x
f_{XY}(x,y) =
0
otherwise.

Calculate P Y < X|X = 13 .



Problem 44.9 ‡
Once a fire is reported to a fire insurance company, the company makes an
initial estimate, X, of the amount it will pay to the claimant for the fire loss.
When the claim is finally settled, the company pays an amount, Y, to the
claimant. The company has determined that $X$ and $Y$ have the joint density
function

2
y −(2x−1)/(x−1) $X$ > 1, y > 1
x 2 (x−1)
f_{XY}(x,y) =
0
otherwise.432
JOINT DISTRIBUTIONS
Given that the initial claim estimated by the company is 2, determine the
probability that the final settlement amount is between 1 and 3 .



Problem 44.10 ‡
A company offers a basic life insurance policy to its employees, as well as a
supplemental life insurance policy. To purchase the supplemental policy, an
employee must first purchase the basic policy.
Let $X$ denote the proportion of employees who purchase the basic policy, and
Y the proportion of employees who purchase the supplemental policy. Let
X and $Y$ have the joint density function f_{XY}(x,y) = 2(x + y) on the region
where the density is positive.
Given that 10% of the employees buy the basic policy, what is the probability
that fewer than 5% buy the supplemental policy?



Problem 44.11 ‡
An auto insurance policy will pay for damage to both the policyholder’s car
and the other driver’s car in the event that the policyholder is responsible
for an accident. The size of the payment for damage to the policyholder’s
car, X, has a marginal density function of 1 for 0 < $X$ < 1. Given $X$ = x, the
size of the payment for damage to the other driver’s car, Y, has conditional
density of 1 for $X$ < y < $X$ + 1.
If the policyholder is responsible for an accident, what is the probability that
the payment for damage to the other driver’s car will be greater than 0.5?



Problem 44.12 ‡
You are given the following information about N, the annual number of
claims for a randomly selected insured:
1
2
1
P(N = 1) =
3
1
P(N > 1) =
6
P(N = 0) =
Let S denote the total annual claim amount for an insured. When N = 1, S
is exponentially distributed with mean 5 . When N > 1, S is exponentially
distributed with mean 8 . Determine P(4 < S < 8).44 CONDITIONAL DISTRIBUTIONS: CONTINUOUS CASE
433



Problem 44.13
Let $Y$ have a uniform distribution on the interval (0, 1), and let the con-
√
ditional distribution of $X$ given $Y = y$ be uniform on the interval (0, y).
What is the marginal density function of $X$ for $0 < X < 1$?



Problem 44.14 ‡
The distribution of Y, given X, is uniform on the interval [0, X]. The marginal
density of $X$ is

2x for 0 < $X$ < 1
f $X$ (x) =
0
otherwise.
Determine the conditional density of X, given $Y = y$ > 0.



Problem 44.15
Suppose that $X$ has a continuous distribution with p.d.f. f $X$ (x) = 2x on
(0, 1) and 0 elsewhere. Suppose that $Y$ is a continuous random variable such
that the conditional distribution of $Y$ given $X$ = $X$ is uniform on the interval
(0, x). Find the mean and variance of Y.



Problem 44.16 ‡
An insurance policy is written to cover a loss $X$ where $X$ has density function
 3 2
x 0 ≤ $X$ ≤ 2
8
f $X$ (x) =
0 otherwise.
The time T (in hours) to process a claim of size x, where 0 ≤ $X$ ≤ 2, is
uniformly distributed on the interval from $X$ to 2x.
Calculate the probability that a randomly chosen claim on this policy is
processed in three hours or more.



Problem 44.17 ‡
A machine has two components and fails when both components fail. The
number of years from now until the first component fails, X, and the number
of years from now until the machine fails, Y, are random variables with joint
density function

1 − (x+y)
e 6
for 0 < $X$ < y
18
f_{XY}(x,y) =
0
otherwise.
Find f Y |X (y|2).434
JOINT DISTRIBUTIONS



Problem 44.18 ‡
As a block of concrete is put under increasing pressure, engineers measure the
pressure $X$ at which the first fracture appears and the pressure $Y$ at which
the second fracture appears. $X$ and $Y$ are measured in tons per square inch
and have joint density function

24x(1 − y) for 0 < $X$ < y < 1
f_{XY}(x,y) =
0
otherwise.
Calculate the conditional density of the pressure at which the second fracture
appears, given that the first fracture appears at 1/3 ton per square inch.


Problem 44.19

f_{XY}(x,y) =
k for 0 < $X$ < y < 1
0
otherwise.
Determine $f_{X|Y}(x|y)$ and f $Y$ |X (y|x).


Problem 44.20
Let $X$ and $Y$ be two continuous random variables with joint density function
f_{XY}(x,y). Show that
$f_{X|Y}(x|y)$f Y (y)
.
f (x|y)f Y (y)dy
−∞ X|Y
f Y |X (y|x) = R ∞


Problem 44.21 ‡
The elapsed time, T, between the occurrence and the reporting of an accident
has probability density function
 8t−t 2
for 0 < t < 6
72
f T (t) =
0
otherwise.
Given that T = t, the elapsed time between the reporting of the accident
and payment by the insurer is uniformly distributed on [2 + t, 10].
Calculate the probability that the elapsed time between the occurrence of
the accident and payment by the insurer is less than 4.